Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-18 Manuscript.Rmd
Update Manuscript.Rmd

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at a landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018]. Furthermore, traditional management regimes are currently changing wich also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

@@

We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3) For more details on the field methods see @Plattner2004, @Roth2013 and @Roth2017. We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria resulted in sample of `r nrow(sites)` sites.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of N deposition see @Rihm2001 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

@@

# Compile res for table
res <- tibble()
tparams <- c("Temperature", "Precipitation", "NTOT", "Inclination")
res <- tibble(
  Gradient = c("Annual mean temperature", "Annual mean precipitation", "Nitrogen deposition", "Inclination"),
  Estimate = fixef(mod)[tparams],
  `5%` = posterior_interval(mod, pars = tparams)[,1],
  `95%` = posterior_interval(mod, pars = tparams)[,2])

@@

In Fig. \@ref(fig:cl) we compare the colonization and local survival probability of oligotrophic (indicator value of nutrients = 2) and eutrophic (indicator value of nutrients = 4) species along the Nitrogen deposition gradient. Local survival probability was the same for oligotrophic and euthrophic species at a deposition rate of `r CL_surv` kg N ha$^{-1}$ yr$^{-1}$; colonization probability was the same for for oligotrophic and euthrophic species at a deposition rate of `r CL_col` kg N ha$^{-1}$ yr$^{-1}$. In only `r mean((sites$NTOT * 10 + 10) < 12.5)`% of the sites the deposition rate was below 12.5 kg N ha$^{-1}$ yr$^{-1}$ where the replacement of eutrophic with oligotrophic species is likely.

@@

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled  Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that mountain hay meadow communities can reach alternative stable states with eutrophic species that are unlikely to disappear again even if Nitrogen depositon is decreasing again [@Stevens2016]. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

@@

Hower, the results of Fig. \@ref(fig:figconsequences) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how species richness is decreasing along the Nitrogen deposition gradient at differnt time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contraticted our prediction that the nitrogen deposition should become weaker over time since Nitrogen deposition decreased. However, we then realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where the colonizating species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in oligotrophic species replacing eutrophic species especially at sites with low deposition, thus leading to a steeper decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition. The increasingly steeper decline over time in the eutrophic species richness along the spatial Nitrogen deposition gradient is thus a sign of the recovery of plant communities in low deposition sites rather than a sign that the Nitrogen deposition effects became stronger over time.

>>>>

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at a landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018](>> Steinbauer<<). Furthermore, traditional management regimes are currently changing wich also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

@@

We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). For more details on the field methods see @Plattner2004, @Roth2013 and @Roth2017. We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria resulted in sample of `r nrow(sites)` sites.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see @Rihm2001 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

@@

# Compile res for table
res <- tibble()
tparams <- c("Temperature", "Precipitation", "NTOT", "Inclination")
res <- tibble(
  Gradient = c("Mean annual temperature", "Mean annual precipitation", "Nitrogen deposition", "Inclination"),
  Estimate = fixef(mod)[tparams],
  `5%` = posterior_interval(mod, pars = tparams)[,1],
  `95%` = posterior_interval(mod, pars = tparams)[,2])

@@

In Fig. \@ref(fig:cl) we compare the colonization and local survival probability of oligotrophic (indicator value of nutrients <3) and eutrophic (indicator value of nutrients > 3) species along the Nitrogen deposition gradient. Local survival probability was the same for oligotrophic and euthrophic species at a deposition rate of `r CL_surv` kg N ha$^{-1}$ yr$^{-1}$; colonization probability was the same for for oligotrophic and euthrophic species at a deposition rate of `r CL_col` kg N ha$^{-1}$ yr$^{-1}$. In only `r mean((sites$NTOT * 10 + 10) < 12.5)`% of the sites the deposition rate was below 12.5 kg N ha$^{-1}$ yr$^{-1}$ where the replacement of eutrophic with oligotrophic species is likely.

@@

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled  Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at higher Nitrogen deposition rates. This suggests that mountain hay meadow communities can reach alternative stable states with eutrophic species that are unlikely to disappear even if Nitrogen depositon is reduced [@Stevens2016]. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

@@

Hower, the results of Fig. \@ref(fig:figconsequences) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how species richness is decreasing along the Nitrogen deposition gradient at differnt time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contraticted our prediction that the Nitrogen deposition should become weaker over time since Nitrogen deposition decreased. However, we then realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where the colonizating species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in oligotrophic species replacing eutrophic species especially at sites with low deposition, thus leading to a steeper decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition. The increasingly steeper decline over time in the eutrophic species richness along the spatial Nitrogen deposition gradient is thus a sign of the recovery of plant communities in low deposition sites rather than a sign that the Nitrogen deposition effects became stronger over time.
